---
title: "Worker Satisfaction Survey — Data De-identification Workbook"
author: "Data De-identification Advisor"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4
)
```

# Introduction

This workbook guides you through de-identifying the **Apple Grower Satisfaction Survey** (`WorkerSatisfaction_300rows.xlsx`: 300 rows, 16 columns).

### Learning objectives

-   Identify privacy risks in survey data
-   Apply de-identification steps in R using **dplyr**
-   Export a shareable dataset and a separate data key file

### Workflow

1.  **Setup** — load packages and set file paths
2.  **Load and assess** — inspect raw data and review privacy risks
3.  **Transform** — apply four de-identification steps (pseudonymization, aggregation, anonymization)
4.  **Verify and deliver** — run QA checks and export output files

### De-identification steps (Part 3)

| Step  | Technique        | Variables                                     |
|-------|------------------|-----------------------------------------------|
| **1** | Pseudonymization | `worker_id`, `email_id`, `owner_id`           |
| **2** | Aggregation      | `age` (date of birth → age bands)             |
| **3** | Pseudonymization | `city`, `orchard_id`                          |
| **4** | Anonymization    | `immigration_stat`, `username_id`, `comments` |

### Terminology

| Technique | Meaning in this workbook |
|----------------------|--------------------------------------------------|
| **Pseudonymization** | Replace identifiers with coded values; a **data key** can reverse the mapping (Steps 1 and 3) |
| **Aggregation** | Group a detailed value into broader categories (Step 2: DOB → age band) |
| **Anonymization** | Irreversibly remove or transform data so individuals cannot be re-identified; **no key file** (Step 4) |

### Output files

-   **`WorkerSatisfaction_300rows_deidentified.xlsx`** — for research (e.g. public access...)
-   **`data_key_file/WorkerSatisfaction_data_key_DUMMY.xlsx`** — for authorized personnel only; store separately from the de-identified data (dummy example)

Do not edit the raw source file (`WorkerSatisfaction_300rows.xlsx`).

### Roadmap

| Part | Section | What happens |
|-------------------|----------------------|-------------------------------|
| **1** | Setup | Install packages, set paths |
| **2** | Load and assess | Read raw data; review risks by de-identification step |
| **3** | Transform | Steps 1–4: pseudonymize → aggregate → pseudonymize → anonymize |
| **4** | Verify and deliver | QA checks → export de-identified file → export data key |

| Step | Technique | What to look for (Part 2) | What to do (Part 3) |
|--------------|--------------|-------------------------|--------------------|
| **1** | Pseudonymization | Names, emails, owner names | Replace with `Worker_01`, `Email_01`, `Owner_01`, … (reversible via key) |
| **2** | Aggregation | `age` is date of birth | Group into 5-year age bands |
| **3** | Pseudonymization | City and orchard names | Replace with `City_01`, `Orchard_01`, … (reversible via key) |
| **4** | Anonymization | Immigration status, handles, free text | Pool small groups; remove `username_id` and `comments` (irreversible) |

------------------------------------------------------------------------

# Part 1 — Setup

> **R and RStudio (workshop prerequisite)**\
> This workbook assumes you have **R** and **RStudio** installed and can open and run `.Rmd` files. If you have not set these up yet, follow the UBC Library Research Commons guide:\
> [Installing R and RStudio](https://ubc-library-rc.github.io/IntroR/content/installation.html)

### Required packages

| Package   | Purpose                             |
|-----------|-------------------------------------|
| `readxl`  | Read the original Excel survey file |
| `dplyr`   | Transform data (native pipe `\|>`)  |
| `writexl` | Export Excel outputs                |

Run **Install packages** once if a package is missing.

```{r install-packages, eval = FALSE}
# install.packages("readxl")
# install.packages("dplyr")
# install.packages("writexl")
# install.packages(c("readxl", "dplyr", "writexl"))
```

```{r packages}
library(readxl)
library(dplyr)
library(writexl)
```

```{r constants}
DATA_FILE   <- "WorkerSatisfaction_300rows.xlsx"
OUTPUT_FILE <- "WorkerSatisfaction_300rows_deidentified.xlsx"
KEY_DIR     <- "data_key_file"
KEY_FILE    <- file.path(KEY_DIR, "WorkerSatisfaction_data_key_DUMMY.xlsx")
K_THRESHOLD <- 5
```

**k-anonymity threshold:** categories with fewer than `r K_THRESHOLD` records will be pooled before anonymization.

```{r show-constants}
cat("K_THRESHOLD =", K_THRESHOLD, "\n")
cat("Data key file:", KEY_FILE, "\n")
```

------------------------------------------------------------------------

# Part 2 — Load and Assess

Load the unmodified survey and review what makes the data identifiable.

## 2.1 Load raw data

```{r load-raw}
raw <- read_excel(DATA_FILE, sheet = "Unmodified data")

cat("Raw data:", nrow(raw), "rows ×", ncol(raw), "columns\n")
head(raw, 10)
```

## 2.2 Step 1 — Direct identifiers (pseudonymization)

These variables point to a specific person. They will be **pseudonymized** in Step 1 (not dropped) so rows stay linkable within the dataset without exposing real names or contact details. A data key allows authorized staff to reverse the mapping.

| Variable    | Why it is risky                       | Pseudonym prefix |
|-------------|---------------------------------------|------------------|
| `worker_id` | Full name (300 unique values)         | `Worker_001`, …  |
| `email_id`  | Email address (300 unique values)     | `Email_001`, …   |
| `owner_id`  | Orchard owner name (12 unique values) | `Owner_01`, …    |

## 2.3 Step 2 — Date of birth (aggregation)

The column `age` stores **date of birth**, not a numeric age. Exact DOB is a strong quasi-identifier when combined with location and employer.

```{r dob-explore}
data.frame(
  earliest     = as.character(min(as.Date(raw$age))),
  latest       = as.character(max(as.Date(raw$age))),
  unique_dates = length(unique(raw$age))
)
```

Exact DOB will be replaced with **5-year age bands** in Step 2 (aggregation).

## 2.4 Step 3 — Geography and orchards (pseudonymization)

`city` has 12 levels; `province` is always B.C. Small groups are easier to re-identify. Orchard names link respondents to a specific workplace. Both will be pseudonymized in Step 3.

```{r geo-explore}
raw |> count(city, sort = TRUE)
```

> **Note:** If any city had fewer than `r K_THRESHOLD` records, we would pool it into `"Other BC community"` before pseudonymizing.

```{r orchard-explore}
raw |> count(orchard_id, sort = TRUE)
```

> **Note:** Replace orchard names with non-descriptive codes (`Orchard_01`, …).

## 2.5 Step 4 — Immigration status, handles, and free text (anonymization)

These variables cannot be safely shared in their original form. **Anonymization** is irreversible — no data key is created for these fields.

### `immigration_stat`

Small category counts increase re-identification risk. Pool any group below the k-anonymity threshold into a broader label.

```{r geo-immigration-check}
raw |> count(immigration_stat, sort = TRUE)
```

```{r immigration-rare}
rare_immigration <- raw |>
  count(immigration_stat) |>
  filter(n < K_THRESHOLD) |>
  pull(immigration_stat)

cat("Rare immigration groups (n <", K_THRESHOLD, "):\n")
rare_immigration
```

> **Note:** In this dataset all immigration groups meet the threshold. The pooling logic in Step 4 still runs as a safeguard.

### `username_id`

Social-media handles are quasi-identifiers (265 unique values; 36 missing). They will be **removed** in Step 4 — anonymization by deletion.

### `comments`

The `comments` field can name people, employers, platforms, and birth years — even after other columns are cleaned.

```{r comments-risk}
raw |>
  filter(comments != "None") |>
  slice(1:10) |>
  select(comments)
```

> **Note:** Remove `comments` entirely in Step 4. Free text cannot be reliably anonymized without manual review.

------------------------------------------------------------------------

# Part 3 — Transform

Apply each step from Part 2 in order. Run chunks sequentially; each step shows **10 rows** of the updated dataframe.

```{r step1-vars}
step1_ids <- c("worker_id", "email_id", "owner_id")
```

## Step 1: Pseudonymization for direct identifiers

Replace `worker_id`, `email_id`, and `owner_id` with non-descriptive codes. Store mappings in the data key (Part 4).

```{r step1-pseudonymize-direct}
step1 <- raw |>
  mutate(
    worker_id = paste0("Worker_", sprintf("%03d", as.integer(factor(worker_id)))),
    email_id  = paste0("Email_",  sprintf("%03d", as.integer(factor(email_id)))),
    owner_id  = paste0("Owner_",  sprintf("%02d", as.integer(factor(owner_id))))
  )

cat("Step 1:", nrow(step1), "rows ×", ncol(step1), "columns\n")
head(step1, 10)
```

```{r step1-id-counts}
step1 |>
  select(all_of(step1_ids)) |>
  slice(1:5)
```

## Step 2: Aggregation for the age variable

Replace exact date of birth with **5-year age bands**.

```{r step2-aggregate-age}
to_age_band <- function(dob) {
  yrs <- as.integer(difftime(Sys.Date(), as.Date(dob), units = "days") / 365.25)
  if (yrs < 25)      "18-24"
  else if (yrs < 35) "25-34"
  else if (yrs < 45) "35-44"
  else if (yrs < 55) "45-54"
  else               "55+"
}

step2 <- step1 |>
  mutate(age_band = sapply(age, to_age_band)) |>
  select(-age) |>
  relocate(age_band, .before = immigration_stat)

cat("Step 2:", nrow(step2), "rows ×", ncol(step2), "columns\n")
head(step2, 10)
```

## Step 3: Pseudonymization for city and orchard_id

Pseudonymize `city` and `orchard_id` with non-descriptive codes.

```{r step3-pseudonymize-city-orchard}
step3 <- step2 |>
  mutate(
    city       = paste0("City_",    sprintf("%02d", as.integer(factor(city)))),
    orchard_id = paste0("Orchard_", sprintf("%02d", as.integer(factor(orchard_id))))
  )

cat("Step 3:", nrow(step3), "rows ×", ncol(step3), "columns\n")
head(step3, 10)
```

```{r step3-city-orchard-counts}
step3 |> count(city, sort = TRUE)
step3 |> count(orchard_id, sort = TRUE)
```

## Step 4: Anonymization for immigration_stat, username_id, and comments

Anonymization is **irreversible** — removed columns and pooled labels are not stored in the data key.

```{r step4-anonymize}
step4 <- step3 |>
  mutate(
    immigration_stat = if_else(
      immigration_stat %in% rare_immigration,
      "Other immigration status",
      immigration_stat
    )
  ) |>
  select(-username_id, -comments)

deid <- step4

cat("Step 4 — final:", nrow(deid), "rows ×", ncol(deid), "columns\n")
head(deid, 10)
```

```{r step4-immigration-counts}
deid |> count(immigration_stat, sort = TRUE)
```

### What changed

| Step | Technique | Variables |
|------------------|---------------------------|---------------------------|
| **1** | Pseudonymization | `worker_id` → `Worker_001` …; `email_id` → `Email_001` …; `owner_id` → `Owner_01` … |
| **2** | Aggregation | `age` (DOB) → `age_band` (5-year groups) |
| **3** | Pseudonymization | `city` → `City_01` …; `orchard_id` → `Orchard_01` … |
| **4** | Anonymization | `immigration_stat` pooled if below threshold; `username_id` and `comments` removed |

------------------------------------------------------------------------

# Part 4 — Verify and Deliver

## 4.1 QA checks

Confirm the de-identified data is safe to export. All checks should show `TRUE`.

```{r qa-checks}
data.frame(
  check = c(
    "Step 1: direct IDs pseudonymized",
    "Step 2: no exact DOB column",
    "Step 3: cities pseudonymized",
    "Step 3: orchards pseudonymized",
    "Step 4: comments removed",
    "Step 4: username_id removed",
    "Row count unchanged"
  ),
  passed = c(
    !any(unique(raw$worker_id) %in% deid$worker_id),
    !"age" %in% names(deid),
    !any(unique(raw$city) %in% deid$city),
    !any(unique(raw$orchard_id) %in% deid$orchard_id),
    !"comments" %in% names(deid),
    !"username_id" %in% names(deid),
    nrow(deid) == nrow(raw)
  )
)
```

```{r qa-checks-step1}
data.frame(
  check = c(
    "Emails pseudonymized",
    "Owners pseudonymized"
  ),
  passed = c(
    !any(unique(raw$email_id) %in% deid$email_id),
    !any(unique(raw$owner_id) %in% deid$owner_id)
  )
)
```

```{r qa-checks-step4}
data.frame(
  check = c(
    "No rare immigration groups in output",
    "age_band present"
  ),
  passed = c(
    all(table(deid$immigration_stat) >= K_THRESHOLD),
    "age_band" %in% names(deid)
  )
)
```

```{r qa-summary-stats}
data.frame(
  rows               = nrow(deid),
  columns            = ncol(deid),
  smallest_city      = min(table(deid$city)),
  smallest_age_band  = min(table(deid$age_band)),
  smallest_orchard   = min(table(deid$orchard_id))
)
```

> **Note:** Each group should ideally have at least `r K_THRESHOLD` records, or be pooled into a broader category.

## 4.2 Export de-identified dataset

This is the file you can share for analysis. It does **not** include the data key.

```{r export}
write_xlsx(deid, OUTPUT_FILE)

cat("Exported:", OUTPUT_FILE, "\n")
cat("Rows:", nrow(deid), "| Columns:", ncol(deid), "\n\n")
cat("Preview of exported data:\n")
head(deid, 10)
```

## 4.3 Data key file (DUMMY — illustrative only)

Keeping the data key file well protected is essential because it is the sole mechanism that can re-identify individuals in otherwise de-identified research data, and its compromise can undermine confidentiality protections and cause harm to participants. In the UBC environment, the key file must be stored separately from research data on UBC-approved secure systems, with access restricted to authorized personnel only and never stored on personal devices or unapproved cloud services.

Here is an example of a data key file for the work described above; this is a **dummy file provided for illustrative purposes only**, and in real-world research settings, such files must never be shared on public systems.

**File path:** `r KEY_FILE` — **one sheet** (`data_key`) with all mappings in a single table:

| Column | Purpose |
|----------------------------------|--------------------------------------|
| `variable` | Which field was pseudonymized (`worker_id`, `email_id`, `city`, …) |
| `original_value` | Value in the raw file |
| `pseudonym_code` | Code used in the de-identified file |
| `notes` | Optional step label |

To **restore** a de-identified file: for each `variable`, join `pseudonym_code` in the de-identified data back to `original_value` using this table (never merge the key into the shareable dataset).

The key holds **mappings for pseudonymized identifiers only** (Steps 1 and 3). Anonymized fields (`username_id`, `comments`) are not included — they cannot be reversed.

```{r build-data-key}
worker_key <- raw |>
  distinct(worker_id) |>
  arrange(worker_id) |>
  transmute(
    original_value = worker_id,
    pseudonym_code = paste0("Worker_", sprintf("%03d", row_number()))
  )

email_key <- raw |>
  distinct(email_id) |>
  arrange(email_id) |>
  transmute(
    original_value = email_id,
    pseudonym_code = paste0("Email_", sprintf("%03d", row_number()))
  )

owner_key <- raw |>
  distinct(owner_id) |>
  arrange(owner_id) |>
  transmute(
    original_value = owner_id,
    pseudonym_code = paste0("Owner_", sprintf("%02d", row_number()))
  )

orchard_key <- raw |>
  distinct(orchard_id) |>
  arrange(orchard_id) |>
  transmute(
    original_value = orchard_id,
    pseudonym_code = paste0("Orchard_", sprintf("%02d", row_number()))
  )

city_key <- raw |>
  distinct(city) |>
  arrange(city) |>
  transmute(
    original_value = city,
    pseudonym_code = paste0("City_", sprintf("%02d", row_number()))
  )

# One table: stack every mapping for simple lookup and restore
data_key <- bind_rows(
  worker_key  |> mutate(variable = "worker_id",  notes = NA_character_),
  email_key   |> mutate(variable = "email_id",   notes = NA_character_),
  owner_key   |> mutate(variable = "owner_id",   notes = NA_character_),
  orchard_key |> mutate(variable = "orchard_id", notes = NA_character_),
  city_key    |> mutate(variable = "city",       notes = NA_character_)
) |>
  select(variable, original_value, pseudonym_code, notes) |>
  arrange(variable, original_value)

cat("Data key preview (first 10 rows):\n")
head(data_key, 10)

cat("\nRows by variable:\n")
data_key |> count(variable)
```

```{r export-data-key}
dir.create(KEY_DIR, showWarnings = FALSE)

write_xlsx(list(data_key = data_key), KEY_FILE)

cat("Data key exported:", KEY_FILE, "\n")
cat("Single sheet: data_key (", nrow(data_key), " mapping rows )\n", sep = "")
cat("Store separately from", OUTPUT_FILE, "— never on public systems.\n")
```

------------------------------------------------------------------------

# Closing Summary

| Step | Technique | What changed |
|-----------------|------------------------|-------------------------------|
| **1** | Pseudonymization | `worker_id`, `email_id`, `owner_id` → pseudonym codes (key file) |
| **2** | Aggregation | `age` (DOB) → `age_band` (5-year groups) |
| **3** | Pseudonymization | `city` → `City_01` …; `orchard_id` → `Orchard_01` … |
| **4** | Anonymization | `immigration_stat` pooled if below threshold; `username_id` and `comments` removed |

The de-identified dataset keeps all satisfaction and treatment ratings for analysis while substantially lowering re-identification risk. **The data key is highly sensitive** — it links pseudonym codes back to real names, emails, and locations for Steps 1 and 3 only. Protect and store it separately.
